---
output: 
  github_document: default
  html_document:
    self_contained: no
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# adjclust: Adjacency-constrained clustering

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/adjclust)](https://cran.r-project.org/package=adjclust)
[![Travis Build Status](https://travis-ci.org/pneuvial/adjclust.svg?branch=master)](https://travis-ci.org/pneuvial/adjclust)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/pneuvial/adjclust?branch=master&svg=true)](https://ci.appveyor.com/project/pneuvial/adjclust)
[![Coverage Status](https://img.shields.io/codecov/c/github/pneuvial/adjclust/master.svg)](https://codecov.io/github/pneuvial/adjclust?branch=master)


`adjclust` is a package that provides methods to perform adjacency-constrained hierarchical agglomerative clustering. Adjacency-constrained hierarchical agglomerative clustering is hierarchical agglomerative clustering (HAC) in which each observation is associated to a position, and the clustering is constrained so as only adjacent clusters are merged. It is useful in bioinformatics (e.g. Genome Wide Association Studies or Hi-C data analysis).

`adjclust` provides three user level functions: `adjClust`, `snpClust` and `hicClust`, which are briefly explained below.

## Installation

You can install adjclust from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("pneuvial/adjclust")
```


## `adjClust` 

`adjClust` performs adjacency-constrained HAC for standard and sparse, similarity and dissimilarity matrices and `dist` objects. `Matrix::dgCMatrix` and `Matrix::dsCMatrix` are the supported sparse matrix classes. Let's look at a basic example

```{r adjClust}
library("adjclust")

sim <- matrix(c(1.0, 0.5, 0.2, 0.1,
                0.5, 1.0, 0.1, 0.2,
                0.2, 0.1, 1.0, 0.6,
                0.1, 0.2 ,0.6 ,1.0), nrow=4)
h <- 3
fit <- adjClust(sim, "similarity", h)
plot(fit)
```

The result is of class `chac`. It can be plotted as a dendogram (as shown above). Successive merge and heights of clustering can be obtained by `fit$merge` and `fit$height` respectively.

## `snpClust` 

`snpClust` performs adjacency-constrained HAC for specific application of Genome Wide Association Studies (GWAS). A minimal example is given below. See [GWAS Vignette](vignettes/snpClust.Rmd) for details.

```{r snpClust}
library("snpStats")
data("ld.example", package = "snpStats")
geno <- ceph.1mb[, -316]  ## drop one SNP leading to one missing LD value
h <- 100
ld.ceph <- ld(geno, stats = "R.squared", depth = h)
image(ld.ceph, lwd = 0)

fit <- snpClust(geno, stats = "R.squared", h = h)
plot(fit)
```

## `hicClust` 

 `hicClust` performs adjacency-constrained HAC for specific application of Hi-C data analysis. A minimal example is given below. See [Hi-C Vignette](vignettes/hicClust.Rmd) for details.

```{r HiTC, message=FALSE}
library("HiTC")
```

```{r hicClust}
data("hic_imr90_40_XX", package = "adjclust")
binned <- binningC(hic_imr90_40_XX, binsize = 5e5)
mapC(binned)

fitB <- hicClust(binned)
plot(fitB)
```


## Credits 

Version 0.4.0 of this package was completed by Shubham Chaturvedi as a part of the [Google Summer of Code 2017](https://summerofcode.withgoogle.com/projects/#4961904920363008) program.

