---
title: "Notes on CHAC implementation in adjclust"
author: "Pierre Neuvial, Nathalie Villa-Vialaneix"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Notes on CHAC implementation in adjclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document aims at clarifying relations between dissimilarity and similarity
methods for hierarchical agglomerative clustering (HAC) and at explaining
implementation choices in `adjclust`. In this document, we suppose given $n$
objects, $\{1, \ldots, n\}$ that have to be clustered using 
adjacency-constrained HAC, that is, in such a way that only adjacent
objects/clusters can be merged.

# Basic implementation of CHAC in `adjclust`

The basic implementation of `adjclust` takes, as an input, a kernel $k$ which
is supposed to be symmetric and positive (in the kernel sense) and to satisfy:
\[
	k(i,i) = 1,\qquad \forall\, i=1,\ldots,n.
\]

If your data are under this format, then the constrained clustering can be 
performed with 
```{r ex-sim, eval=FALSE}
fit <- adjClust(k, type = "similarity")
```
or with
```{r ex-sim2, eval=FALSE}
fit <- adjClust(k, type = "similarity", h = h)
```
if, in addition, the kernel $k$ is supposed to have only null entries outside
of a diagonal of size `h`.

The implementation is the one described in [1].

# More advanced used for kernel or similarity matrices

## Non positive but normalized similarities

In this section, the available data set is a matrix $s$ that can either have 
only positive entries (in this case it is called a similarity) or both positive 
and non-positive entries. If, in addition, $s(i,i) = 1$ for all $i=1,\ldots,n$ 
and $s(i,j) \leq 1$ for all $i,j=1,\ldots,n$ then the algorithm implemented in
`adjclust` can be applied directly, similarly as for a standard kernel
(section 1). This section explains why this is the case.

The interpretation is similar to the kernel case, under the assumption that 
small similarity values or similarity values that are strongly negative are less 
expected to be clustered together than large similarity values. The application
of the method is justified by the fact that, for a given matrix $s$ described
as above, we can find a $\lambda > 0$ such that the matrix $k_\lambda$ defined
by
\[
  \forall\,1,\ldots,n,\qquad k_\lambda(i,j) = s(i,j) + \lambda 
  \mathbb{1}_{\{i=j\}}
\]
is a kernel (*i.e.*, the matrix $k = s + \lambda I$ is positive 
definite; indeed, it is the case for any $\lambda$ larger than the opposite of 
the smallest negative eigenvalue of $s$. [2] shows that the HAC obtained from 
the distance induced by the kernel $k_\lambda$ in its feature space and the HAC
obtained from the *ad hoc* dissimilarity defined by
\[
  \forall\, i,j=1,\ldots,n,\qquad d(i,j) = \sqrt{s(i,i) + s(j,j) - 2s(i,j)} = 
  \sqrt{2(1-s(i,j))}
\]
are identical, except that all the merging levels are shifted by $\lambda$. 

In conclusion, to address this case, the command lines that have to be used are
the ones described in section 1.

## Non normalized similarities

Suppose now that the data set is described by a matrix $s$ as in the previous
section except that which does not satisfy one of the two following conditions:

* [C1] $\forall i \in\{1,\ldots,n\}$, $s(i,i)=1$
* [C2] $\forall i,j \in\{1,\ldots,n\}$,
\[
  2s(i,j) \leq s(i,i) + s(j,j).
\]

These properties were implicitly assumed to hold in the original implementation 
of constrained HAC described in [1]. Specifically, this implementation relied on 
the implicit dissimilarity
\[
  \forall\,i,j=1,\ldots,n,\qquad d^2(i,j) = s(i,i) + s(j,j) - 2s(i,j)
\]
which has no meaning if 2. above does not hold. If either [C1] or [C2] is not
met, the package performs the following pre-transformations:

1. a matrix $s^{*}$ is defined as
\[
  \forall\,i,j=1,\ldots,n,\qquad s^{*}(i,j) = s(i,j) + \lambda 
  \mathbb{1}_{\{i=j\}}
\]
for a $\lambda$ large enough to ensure that [C2] holds for $s^{*}$. In the 
package, $\lambda$ is chosen as
\[
  \lambda := \epsilon + \max_{i,j} \left(s(i,j) - s(i,i)\right)_+
\]
for a small $\epsilon > 0$. This case is justified by the property described in 
Section 2.1 (Non-positive but normalized similarities). The underlying idea is
that, shifting the diagonal entries of a similarity matrix does not change HAC
result and thus they can be shifted until they induce a proper *ad-hoc* 
dissimilarity matrix;

1. $s^{*}$ is then normalized to unit diagonal ([C1]), in order to allow its use
in the original implementation as in [1]:
\[
  \forall\,i,j=1,\ldots,n,\qquad s^{**} =s^{*}(i,j) + 1 - 
  \frac{1}{2} (s^{*}(i,i) +s^{*}(j,j)).
\]


By definition, $s^{**}$ satisfies [C1] and [C2], and the *ad-hoc* 
dissimilarities induced by $s^{*}$ and  $s^{**}$ are identical:
\[
  \forall\,i=1,\ldots,j,\qquad d^{**}(i,j) = d^{*}(i,j).
\]

**Important remark**: The transformation described above affects all the 
elements of the similarity matrix (and not only its diagonal), contrarily to 
what happens in section 2.1. Hence, it is not (yet) compatible with the fast 
implementation that requires to have a $h$-band centered on the diagonal outside
of which all entries are supposed to be 0. The current behavior of the package
consists in forbidding the use of the fast implementation when the provided
matrix is not normalized (all diagonal entries equal to 1). For all other cases,
the constrained clustering from $s$, including the above-described matrix
transformations, is available with the full matrix, i.e.:

```{r ex-notnormsim, eval=FALSE}
fit <- adjClust(s, type = "similarity")
```

## Case of dissimilarity data

The original implementation of (unconstrained) HAC in `stats::hclust` takes as 
input a dissimilarity matrix. However, the implementation of `adjclust` is based
on a kernel/similarity approach. We describe in this section how the case of 
dissimilarity is handled.

Suppose given a dissimilarity $d$ which satisfies:

* $d$ has non negative entries: $d(i,j) \geq 0$ for all $i=1,\ldots,n$;

* $d$ is symmetric: $d(i,j) = d(j,i)$ for all $i,j=1,\ldots,n$;

* $d$ has a null diagonal: $d(i,i) = 0$ for all $i=1,\ldots,n$.

Any sequence of positive numbers $(a_i)_{i=1,\ldots,n}$ would provide a 
similarity $s$ for which $d$ is the *ad-hoc* dissimilarity by setting:
\[
  \left\{ \begin{array}{l}
    s(i,i) = a_i\\
    s(i,j) = \frac{1}{2} (a_i + a_j - d^2(i,j))
  \end{array} \right. .
\]
By definition, such an $s$ satisfies [C2], and any choice for 
$(a_i)_{i=1,\ldots,n}$ yields the same clustering (since they all correspond to 
the same *ad-hoc* dissimilarity). To satisfy [C1], the choice $a_i = 1$ for all
$i=1,\ldots,n$ has been made.

The similarity $s$ is thus processed as described in the previous sections. In
short, since by definition it satisfies [C1] and [C2] it is processed as 
described in Sections 1 or 2.1 and the basic and the sparse implementations
are both available with, respectively,

```{r ex-dissim, eval=FALSE}
fit <- adjClust(d, type = "dissimilarity")
```
and
```{r ex-dissim-sparse, eval=FALSE}
fit <- adjClust(d, type = "dissimilarity", h = h)
```


# References

[1] Dehman A. (2015). 
[Spatial clustering of linkage disequilibrium blocks for genome-wide association studies](https://tel.archives-ouvertes.fr/tel-01288568/). 
Phd Thesis, Université Paris Saclay.

[2] Miyamoto S., Abe R., Endo Y., Takeshita J. (2015) Ward method of 
hierarchical clustering for non-Eclidean similarity measures. In: *Proceedings
of the VIIth International Conference of Soft Computing and Pattern 
Recognition* (SoCPaR 2015).
